Understanding Implications of Dataset Choice for Feature Effect Estimation: A Simulation-Based Investigation through Error Decomposition

### 2025-02-20 paper/paper.tex
paper: refine chapters 5 & 6 and appendix

\subsection{Main Experiment}

We give the results of the main experiment error decomposition for the
SimpleNormalCorrelated dataset in \textbf{Fig.\@~\ref{fig:pdp-results-snc}} for
PDP and \textbf{Fig.\@~\ref{fig:ale-results-snc}} for ALE. The full results for  % chktex 13
all datasets, models, and features can be found in
\textsc{Appendix~\ref{app:additional-results}}.

Generally, we observe that the feature effect error is in the most cases
highest when the effect is estimated on validation data. For ALE, this
observation is more pronounced than for PDP. This increase in error is mainly  % chktex 13
due to an increase in variance (see e.g.,
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}} or
\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}). For many models, the
feature effect error is lowest when estimated in a cross-validated manner. This
is slightly more pronounced for overfitting models (OF) than for optimally
tuned models (OT), and is mainly due to a decrease in variance (see e.g.,
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}} and
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance composes
of the model variance and the Monte Carlo variance, both could be reduced by
cross-validation. Monte Carlo variance may decrease as we have slightly more
samples overall compared to the training set. Conversely, the model variance
may decrease as it is averaged out through the cross-validation procedure,
which may especially improve overfitting models that have a higher model
variance. We further investigate these hypotheses in
\textsc{Section~\ref{sec:results-variance-decomposition}}.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_GAM_OF.png}
        \caption{GAM\_OF}
        \label{fig:pdp-results-snc-gam-of}  % chktex 24
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_GAM_OT.png}
        \caption{GAM\_OT}
        \label{fig:pdp-results-snc-gam-ot}  % chktex 24
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_XGBoost_OF.png}
        \caption{XGBoost\_OF}
        \label{fig:pdp-results-snc-xgboost-of}  % chktex 24
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_XGBoost_OT.png}
        \caption{XGBoost\_OT}
        \label{fig:pdp-results-snc-xgboost-ot}  % chktex 24
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_SVM_OT.png}
        \caption{SVM\_OT}
        \label{fig:pdp-results-snc-svm-ot}  % chktex 24
    \end{subfigure}
    \caption{Main experiment results for PDP on SimpleNormalCorrelated dataset.\dots}
    \label{fig:pdp-results-snc}  % chktex 24
\end{figure}

Interestingly, for ALE, we observe for some cases an increased bias when
estimating the feature effects on validation data and via cross-validation
compared to an estimation on training data (see e.g.,
\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-ot}} and
\textbf{Fig.\@~\ref{fig:ale-results-snc-svm-ot}}). Note that this is only the
case when errors are generally low, and the bias is still small in absolute
terms. Since this only occurs for ALE and only for the smaller dataset size
(1000 training samples, 250 validation samples), we hypothesize that this is
due to our approach of choosing the intervals based on theoretical quantiles.
For the smaller validation sets, we may have intervals with few or no samples,
leading to a bias in the estimation of the effect. This hypothesis is supported
by the fact that the increase in bias only occurs in regions with low sample
density, which we demonstrate in
\textsc{Appendix~\ref{app:main-study-further-investigation}}. Thus, an open
question remains whether choosing the intervals based on the empirical
quantiles mitigates this issue.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_GAM_OF.png}
        \caption{GAM\_OF}
        \label{fig:ale-results-snc-gam-of}  % chktex 24
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_GAM_OT.png}
        \caption{GAM\_OT}
        \label{fig:ale-results-snc-gam-ot}  % chktex 24
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_XGBoost_OF.png}
        \caption{XGBoost\_OF}
        \label{fig:ale-results-snc-xgboost-of}  % chktex 24
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_XGBoost_OT.png}
        \caption{XGBoost\_OT}
        \label{fig:ale-results-snc-xgboost-ot}  % chktex 24
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_SVM_OT.png}
        \caption{SVM\_OT}
        \label{fig:ale-results-snc-svm-ot}  % chktex 24
    \end{subfigure}
    \caption{Main experiment results for ALE on SimpleNormalCorrelated dataset.\dots}
    \label{fig:ale-results-snc}  % chktex 24
\end{figure}

Overall, we find no evidence against the currently dominant practice of
estimating feature effects on training data. Further results on Friedman1 and
Feynman I.29.16, as well as for the dummy features, largely confirm our
described findings. For details, we refer to
\textsc{Appendix~\ref{app:additional-results}}. This also means that we observe
no substantial differences between scenarios with correlated features
(SimpleNormalCorrelated) and independent features (Friedman1, Feynman I.29.16).
Similarly, the more real-world relevant Feynman I.29.16 dataset shows results
in line with the other datasets. Note that we only observe one case where the
error of ALE is substantially larger when estimated on training data due to a
high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset,
which we further look into in
\textsc{Appendix~\ref{app:main-study-further-investigation}}.

@@

We use the variance results from the main experiment and the Monte Carlo
variance estimated in the ablation study to show the decomposition of the
feature effect variance into model variance and Monte Carlo variance. The
results for the Friedman1 dataset are given in
\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}} for PDP and
\textbf{Fig.\@~\ref{fig:ale-variance-decomp}} for ALE. Note that we now  % chktex 13
consider the Friedman1 dataset as it provides a wider variety of different
feature effects. The full results including the other datasets are provided in
\textsc{Appendix~\ref{app:additional-variance-decomposition}}.

We observe that the Monte Carlo variance is generally highest when the feature
effects are estimated on the smaller validation set, which aligns with our
expectations. Computing the feature effects in a cross-validated manner leads
to the lowest Monte Carlo variance with a substantial reduction compared to
estimation solely on training or validation data.

In addition, we observe that the model variance is also reduced in many cases when
estimating the feature effects in a cross-validated manner compared to an estimation
on training or validation data, where model variance is (roughly) the same. The observed
reduction through cross-validation is especially
pronounced for the overfitting models (OF) and best visible for PDP
(\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}}). These observations are in line
with our previous hypothesis that cross-validation reduces both model variance by averaging over
different models as well as Monte Carlo variance by averaging over different
samples. This is especially beneficial for overfitting models, which have a
higher model variance.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_pdp_XGBoost_OF.png}
        \caption{XGBoost\_OF PDP on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_pdp_XGBoost_OT.png}
        \caption{XGBoost\_OT PDP on Friedman1}
    \end{subfigure}
    \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
        main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
        obtained by subtracting the Monte Carlo variance from the total variance. The second linear feature
        ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
        on 8000 training samples were not obtained due to excessive computation time.}
    \label{fig:pdp-variance-decomp}  % chktex 24
\end{figure}

While the variance due to MC integration is only a minor component of the total
variance for PDP, it is the dominant component for ALE. Only for feature effect  % chktex 13
estimation via cross-validation, the Monte Carlo variance is again reduced to a
negligible level. This suggests that ALE is more sensitive to the dataset size
than PDP, which we further investigate in
\textsc{Section~\ref{sec:results-dataset-size}}. We also suspect a certain
amount of inaccuracy in the Monte Carlo variance estimation for ALE, which (a)
leads to differences in the model variance between training and validation set
and (b) causes the model variance to become negative in some cases (see
\textbf{Fig.\@\ref{fig:ale-variance-decomp}}). Naturally, this is not possible
and can most likely be attributed to numerical inaccuracies in the MC variance
estimation.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_ale_XGBoost_OF.png}
        \caption{XGBoost\_OF ALE on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_ale_XGBoost_OT.png}
        \caption{XGBoost\_OT ALE on Friedman1}
    \end{subfigure}
    \caption{Variance decomposition results for ALE on Friedman1 dataset. Variance is taken from the
        main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
        obtained by subtracting the Monte Carlo variance from the total variance. The second linear feature
        ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
        on 8000 training samples were not obtained due to excessive computation time. Negative variances are
        due to numerical inaccuracies, marked by the hatch pattern.}
    \label{fig:ale-variance-decomp}  % chktex 24
\end{figure}

For both PDP and ALE, we find that the Monte Carlo variance, as expected,
reduces with larger dataset size. This becomes clear both when comparing
training and validation data and when comparing the two dataset sizes (1000 and
8000 samples). To analyze this not only for those few dataset sizes, we
dedicate a separate ablation study to investigate this further in the next
section.

@@

We show the effect of the dataset size on the Monte Carlo variance for the
features of the Friedman1 dataset for PDP and ALE in
\textbf{Fig.\@~\ref{fig:mc-variance}}. The full results are provided in
\textsc{Appendix~\ref{app:additional-mc-results}}. As expected, we observe that
the Monte Carlo variance decreases with increasing dataset size (except for
$x_3$ and $x_4$ for PDP, where the variance is generally very small). We
find differences between the features: for more complex features like $x_1$ and
$x_2$, which include interactions, the Monte Carlo variance is generally higher
compared to simpler features like $x_3$ and $x_4$. Nonetheless, we observe an
interesting difference between PDP and ALE. While they converge to a similar  % chktex 13
Monte Carlo variance for large dataset sizes, the initial Monte Carlo variance
for very small sample sizes is substantially higher for ALE than for PDP. This  % chktex 13
demonstrates that ALE is more sensitive to the dataset size than PDP. This is  % chktex 13
most likely due to the fact that ALE estimates effects per interval and thus
requires more samples to estimate the effect accurately compared to PDP, which
uses the full dataset to estimate the effect. Note that this analysis
also proves that the Monte Carlo variance introduced by estimating the ground truth
feature effects in the main study is for most cases negligible compared to the
other error sources.

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/mean_mc_variance_friedman1_pdp.png}
        \caption{Mean Monte Carlo variance for PDP}
        \label{fig:mc-variance-pdp}  % chktex 24
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/mean_mc_variance_friedman1_ale.png}
        \caption{Mean Monte Carlo variance for ALE}
        \label{fig:mc-variance-ale}  % chktex 24
    \end{subfigure}
    \caption{Mean Monte Carlo variance for PDP and ALE on Friedman1 dataset.
        For each data size $n_{mc}$, variances are averaged over all grid points.
        Second linear feature ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability.}
    \label{fig:mc-variance}  % chktex 24
\end{figure*}

In addition, a detailed analysis of the Monte Carlo variance per grid point
reveals that for PDP, the variance mostly depends on the sample density, while
for ALE, the variance is also influenced by the shape of the underlying
function. We provide detailed results in
\textsc{Appendix~\ref{app:additional-mc-results}}.

\clearpage
\section{Conclusions}\label{sec:conclusion}

@@

Summarizing our main contributions, we (1) empirically investigated the
impact of dataset choice for feature effect estimation. We (2) based our study on
a theoretical framework that decomposes the feature effect error into different
components and introduces estimators for each of them. We (3) conducted a
review of test functions for simulation studies and build the choices for
our study on these.

Our comprehensive simulation study provided several key insights into
the choice of dataset for feature effect estimation. First and foremost,
we find no general evidence that speaks against the current predominant practice
of using training data for computing feature effects. This is an important validation
of existing approaches in the literature.

More specifically, our results demonstrated that feature effect errors are
generally lower when more data is available for the estimation, which is usually
the case for the training set. This is primarily due to reduced Monte Carlo variance,
as shown in our ablation studies. The effect is particularly pronounced for ALE plots,
which exhibit higher sensitivity to dataset size compared to PDPs, especially for smaller samples,
due to the interval-based estimation. This suggests that for ALE plots, it is even more
important to use a sufficently large dataset for accurate estimation. Based on these
findings, we generally recommend using the largest available dataset for feature effect
estimation, as long as this is computationally feasible.

An interesting finding emerged regarding cross-validation for feature effect estimation.
Our main study suggests that cross-validation is preferable in many scenarios,
leading to lower overall errors. The ablation study reveals that this improvement stems from two distinct mechanisms:
(1) reduced Monte Carlo variance, and (2) reduced model variance through the averaging of estimates from different models.
This dual benefit is particularly advantageous for overfitting models, which typically suffer
from higher variance. Therefore, we recommend using cross-validation for feature effect estimation
when overfitting is a serious concern (e.g., due to a large difference between training
and validation performance). Estimating feature effects for a single model and dataset
or using cross-validation also becomes a trade-off between computation time and accurate
feature effect estimation, as cross-validation comes with computational overhead depending
on the number of folds.

We also observed that the correlation structure between features appears to have limited impact on our findings
and are largely consistent across our set of different, diverse datasets.
This suggests that our recommendations are robust across different data scenarios,
including those with complex feature interactions, correlations, 
and real-world inspired relationships as in the Feynman dataset.

@@

While our study provides valuable insights, several limitations and opportunities
for future research remain. One methodological limitation lies in our choice of using
theoretical quantiles for the grid points in feature effect estimation to make feature
effects comparable. However, this can lead to empty or sparsely populated intervals, particularly for smaller
datasets or in regions of low data density. Moreover, this approach has limited real-world
applicability, as we usually do not now the theoretical but only the empirical quantiles
(which should be the same in expectation).

Furthermore, the observed instabilities in variance estimation for the Monte Carlo variance,
particularly of ALE plots, indicate room for methodological improvements. Future research may
focus on developing more robust estimation procedures for decomposing variance components,
before applying these estimators to address other research questions.

@@

Further, it is important to note that our study primarily adopts a ``true to the data'' perspective
when defining feature effect errors, focusing on explaining the underlying data-generating process
rather than the model's internal representation. While this choice aligns with many practical
applications, there are also use cases for which ``true to the model'' interpretations might be 
preferred, which is not considered in our analysis.
% also: data choice only with regard to feature effect error, not uncertainty estimation

Future research could also investigate the impact of data shifts between training and validation sets.
While our focused considered overfitting models, future work could explore the impact of data shifts
between training and validation sets and their impact on feature effect estimation, analyzing the
robustness of the estimation under distribution shifts.

Moreover, while our cross-validation strategy showed promising improvements over
conventional estimation on a single dataset, this approach could be enhanced further.
Future research could explore strategies that utilize more data per fold, such as using the
training set in each fold for feature effect estimation (and use the validation part only for
perfomance estimation) or combining training and validation splits. Alternatively, bootstrapping
strategies similar to the approaches pursued by Molnar et al.~\cite{molnar_relating_2023} could be
explored.

Following the observations of Loecher~\cite{loecher_debiasing_2024} that hat MDI and SHAP show
a bias toward high-entropy variables for Random Forests, future work might also investigate if and how
feature entropy and cardinality influence feature effect estimation.

While we aimed to incorporate real-world relevance through physics-inspired datasets, further
work is needed to fully capture the complexity of practical applications. In practice, datasets
have features with varying and also just partial effect strengths, different noise levels,
higher-order interactions, and diverse marginal distributions. Future work could focus on
developing more realistic datasets capturing these aspects, to further bridge the gap between
research, such as simulation studies, and practical applications.

These limitations and future directions highlight the rich opportunities for further research
in this area to obtain practically relevant insights.

@@

>>>>

\subsection{Main Experiment Results}

We present the results of the main experiment error decomposition for the GAM and
XGBoost models on the
SimpleNormalCorrelated dataset in \textbf{Fig.\@~\ref{fig:pdp-results-snc}} for
PDP and \textbf{Fig.\@~\ref{fig:ale-results-snc}} for ALE. The full results for  % chktex 13
all datasets, models, and features can be found in
\textsc{Appendix~\ref{app:additional-results}}.

Our analysis reveals that feature effect error is generally highest when the effect
is estimated on validation data. For ALE, this
observation is more pronounced than for PDP. The increase in error is mainly  % chktex 13
due to an increase in variance (see e.g.,
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}} or
\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}). For many models, the
feature effect error is lowest when effects are estimated in a cross-validated manner. This
is slightly more pronounced for overfitting models (OF) than for optimally
tuned models (OT), and mainly stems from a reduced variance (see e.g.,
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}} and
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance is composed
of model variance and Monte Carlo variance, cross-validation could potentially
reduce each of these components. Monte Carlo variance may decrease due to
the slightly larger effective sample size compared to using only the training set.
Conversely, model variance
may decrease as cross-validation averages out estimation variability across multiple
models, which is particularly beneficial for overfitting models that typically exhibit
higher model variance. We further investigate these hypotheses in
\textsc{Section~\ref{sec:results-variance-decomposition}}.

\begin{figure}[ht!]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_GAM_OF.png}
        \caption{GAM\_OF}
        \label{fig:pdp-results-snc-gam-of}  % chktex 24
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_GAM_OT.png}
        \caption{GAM\_OT}
        \label{fig:pdp-results-snc-gam-ot}  % chktex 24
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_XGBoost_OF.png}
        \caption{XGBoost\_OF}
        \label{fig:pdp-results-snc-xgboost-of}  % chktex 24
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_XGBoost_OT.png}
        \caption{XGBoost\_OT}
        \label{fig:pdp-results-snc-xgboost-ot}  % chktex 24
    \end{subfigure}
    \caption{Main experiment results for PDP on SimpleNormalCorrelated dataset for GAM and
        XGBoost models. The other models as well as the dummy features $x_3$ and $x_4$
        are omitted for brevity.}
    \label{fig:pdp-results-snc}  % chktex 24
\end{figure}

\noindent Interestingly, for ALE, we observe for some cases an increased bias when
estimating the feature effects on validation data or through cross-validation
compared to an estimation on training data (see e.g.,
\textbf{Fig.\@~\ref{fig:ale-results-snc-gam-ot}} and
\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-ot}}). However, this phenomenon
occurs only when errors are generally low, and the bias is still small in absolute
terms. Since this appears exclusively for ALE and only with smaller dataset sizes
(1000 training samples, 250 validation samples), we hypothesize it stems from our
approach of defining intervals based on theoretical quantiles.
For smaller validation sets, this may result in intervals with few or no samples,
introducing a bias in the estimation of the effect. This hypothesis is supported
by the fact that the increased bias only occurs in regions with low sample
density, which we demonstrate in
\textsc{Appendix~\ref{app:main-study-further-investigation}}. Thus, an open
question remains whether the practical approach of choosing intervals based on
empirical quantiles would mitigate this issue.

\begin{figure}[ht!]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_GAM_OF.png}
        \caption{GAM\_OF}
        \label{fig:ale-results-snc-gam-of}  % chktex 24
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_GAM_OT.png}
        \caption{GAM\_OT}
        \label{fig:ale-results-snc-gam-ot}  % chktex 24
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_XGBoost_OF.png}
        \caption{XGBoost\_OF}
        \label{fig:ale-results-snc-xgboost-of}  % chktex 24
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_XGBoost_OT.png}
        \caption{XGBoost\_OT}
        \label{fig:ale-results-snc-xgboost-ot}  % chktex 24
    \end{subfigure}
    \caption{Main experiment results for ALE on SimpleNormalCorrelated dataset for GAM and
    XGBoost models. The other models as well as the dummy features $x_3$ and $x_4$
    are omitted for brevity.}
    \label{fig:ale-results-snc}  % chktex 24
\end{figure}

Overall, we find no evidence against the currently dominant practice of
estimating feature effects on training data. Further results on Friedman1 and
Feynman I.29.16, as well as for the dummy features and the SVM largely confirm our
described findings. For details, we refer to
\textsc{Appendix~\ref{app:additional-results}}. This implies that we observe
no substantial differences between scenarios with correlated features
(SimpleNormalCorrelated) and independent features (Friedman1, Feynman I.29.16).
Similarly, the more real-world relevant Feynman I.29.16 dataset produces results
in line with the other datasets. Note that we only observe one case where the
error of ALE is substantially larger when estimated on training data due to a
high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset,
which we further examine in
\textsc{Appendix~\ref{app:main-study-further-investigation}}.

@@

Using variance results from the main experiment and Monte Carlo variance estimates
from our ablation study, we decompose feature effect variance into its model
variance and Monte Carlo variance components.
Results for the Friedman1 dataset are presented in
\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}} for PDP and
\textbf{Fig.\@~\ref{fig:ale-variance-decomp}} for ALE. Note that we now  % chktex 13
focus on Friedman1 as it offers a wider range of different
feature effects. Full results including the other datasets are provided in
\textsc{Appendix~\ref{app:additional-variance-decomposition}}.

Our analysis reveals that Monte Carlo variance is consistently highest when feature
effects are estimated on the smaller validation set, which aligns with our
expectations. Cross-validated estimation yields the lowest Monte Carlo variance with
a substantial reduction compared to
estimation on either training or validation data alone.

Additionally, model variance is also reduced in many cases through cross-validation
compared to estimation on training or validation data, where model variance is
approximately equal. The observed reduction through cross-validation is especially
pronounced for the overfitting models (OF) and best visible for PDP
(\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}}). These observations are in line
with our previous hypotheses that cross-validation reduces both model variance
(by averaging across different models) and Monte Carlo variance (by leveraging
different samples). This is particularly beneficial for overfitting models, which
typically exhibit higher model variance.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_pdp_XGBoost_OF.png}
        \caption{XGBoost\_OF PDP on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_pdp_XGBoost_OT.png}
        \caption{XGBoost\_OT PDP on Friedman1}
    \end{subfigure}
    \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
        main experiment, Monte Carlo variance is estimated in the ablation study. Model variance is
        obtained by subtracting Monte Carlo variance from total variance. The second linear feature
        ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
        on 8000 training samples were not obtained due to excessive computation time.}
    \label{fig:pdp-variance-decomp}  % chktex 24
\end{figure}

\noindent We observe a striking difference between PDP and ALE:\@ while Monte Carlo
variance represents only a minor component of total variance for PDP, it dominates
the variance composition for ALE. Only with cross-validated estimation,  % chktex 13
ALE's Monte Carlo variance is reduces to a negligible level. This suggests that
ALE is more sensitive to the dataset size
than PDP, which we further investigate in
\textsc{Section~\ref{sec:results-dataset-size}}. We note some potential inaccuracies
in Monte Carlo variance estimation for ALE, which manifest as: (a) differences
in model variance between training and validation sets, and (b) negative model
variance values in certain cases (see
\textbf{Fig.\@\ref{fig:ale-variance-decomp}}). Since negative variance is theoretically
impossible, we attribute these artifacts to numerical imprecisions in Monte Carlo variance
estimation.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_ale_XGBoost_OF.png}
        \caption{XGBoost\_OF ALE on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_ale_XGBoost_OT.png}
        \caption{XGBoost\_OT ALE on Friedman1}
    \end{subfigure}
    \caption{Variance decomposition results for ALE on Friedman1 dataset. Variance is taken from the
        main experiment, Monte Carlo variance is estimated in the ablation study. Model variance is
        obtained by subtracting Monte Carlo variance from total variance. The second linear feature
        ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
        on 8000 training samples were not obtained due to excessive computation time. Negative variances, marked
        by the hatch pattern, are due to numerical inaccuracies.}
    \label{fig:ale-variance-decomp}  % chktex 24
\end{figure}

\noindent For both PDP and ALE, we find that the Monte Carlo variance, as expected,
reduces with larger dataset size. This becomes clear both when contrasting
training and validation data and when comparing the two dataset sizes (1000 and
8000 samples). We systematically analyze this relationship across a broader range
of dataset sizes in the next section.

@@

We show the effect of dataset size on Monte Carlo variance for four features
of the Friedman1 dataset for PDP and ALE in
\textbf{Fig.\@~\ref{fig:mc-variance}}. Complete results are provided in
\textsc{Appendix~\ref{app:additional-mc-results}}. As expected, we observe that
Monte Carlo variance generally decreases with increasing dataset size, with exceptions
for PDP on features $x_3$ and $x_4$ where variance remains consistently low across all sample sizes.
We observe feature-specific patterns: more complex features with interactions ($x_1$ and $x_2$)
exhibit higher Monte Carlo variance compared to simpler features ($x_3$ and $x_4$).

A striking difference emerges between PDP and ALE estimation methods. While both converge
to similar Monte Carlo variance levels with large datasets, ALE exhibits substantially
higher variance with small sample sizes. This
demonstrates that ALE is more sensitive to dataset size compared to PDP. This is  % chktex 13
most likely due to the fact that ALE estimates effects per interval and thus
requires more samples to estimate the effect accurately, whereas PDP
uses the full dataset to estimate the effect per grid point.

This analysis also confirms that the Monte Carlo variance introduced by our ground truth
feature effect estimation in the main study is negligible compared to other error sources
for most scenarios.

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/mean_mc_variance_friedman1_pdp.png}
        \caption{Mean Monte Carlo variance for PDP}
        \label{fig:mc-variance-pdp}  % chktex 24
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/mean_mc_variance_friedman1_ale.png}
        \caption{Mean Monte Carlo variance for ALE}
        \label{fig:mc-variance-ale}  % chktex 24
    \end{subfigure}
    \caption{Mean Monte Carlo variance for PDP and ALE on Friedman1 dataset.
        For each data size $n_{mc}$, variances are averaged over all grid points.
        The second linear feature ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability.}
    \label{fig:mc-variance}  % chktex 24
\end{figure*}

\noindent Further analysis of the Monte Carlo variance per grid point
reveals distinct patterns: for PDP, Monte Carlo variance is
primarily influenced by sample density and shows larger values
at distribution edges, while ALE variance mostly depends on 
the underlying function's shape. We provide detailed grid point-level
results in \textsc{Appendix~\ref{app:additional-mc-results}}.

\section{Conclusions}\label{sec:conclusion}

@@

Summarizing our main contributions, we (1) empirically investigated the impact of
dataset choice for feature effect estimation, (2) embedded in a theoretical framework
for decomposing feature effect error into distinct components with corresponding estimators,
and (3) conducted a review of test functions for simulation studies to inform our experimental design.

Our comprehensive simulation study provided several key insights regarding 
the choice of dataset for feature effect estimation. First and foremost,
we found no general evidence that speaks against the current predominant practice
of using training data for estimating feature effects. This is an important validation
of existing approaches.

More specifically, our results demonstrated that feature effect errors are
generally lower when more data is available for estimation, which is usually
the case for the training set. This is primarily due to reduced Monte Carlo variance,
as shown in our ablation studies. The effect is particularly pronounced for ALE plots,
which exhibit higher sensitivity to dataset size compared to PDPs, especially for smaller sample
sizes, due to interval-based estimation. This suggests that for ALE plots, it is even more
important to use a sufficently large dataset for accurate estimation. Based on these
findings, we generally recommend using the largest available dataset for feature effect
estimation, as long as this is computationally feasible.

An interesting finding emerged regarding cross-validation for feature effect estimation.
Our main study results suggest that cross-validation is preferable in many scenarios,
leading to lower overall errors. The ablation study revealed that this improvement stems from two distinct mechanisms:
(1) reduced Monte Carlo variance, and (2) reduced model variance through the averaging of estimates from different models.
This dual benefit is particularly advantageous for overfitting models, which typically suffer
from higher variance. Therefore, we recommend using cross-validation for feature effect estimation
when overfitting is a serious concern (e.g., due to a large difference between training
and validation performance). Estimating feature effects for a single model and dataset
or using cross-validation also becomes a trade-off between computation time and accurate
feature effect estimation, as cross-validation comes with computational overhead depending
on the number of folds.

We also observed that the correlation structure between features appears to have limited impact on our findings,
which are largely consistent across our set of different, diverse datasets.
This suggests that our recommendations are robust across different data scenarios,
including those with complex feature interactions, correlations, 
and real-world inspired relationships as in the Feynman dataset.

@@

While our study provides valuable insights, several limitations and opportunities
for future research remain. One methodological limitation lies in our choice of using
theoretical quantiles for the grid points in feature effect estimation to make feature
effects comparable. However, this can lead to empty or sparsely populated intervals, particularly for smaller
datasets or in regions of low data density. Moreover, this approach has limited real-world
applicability, as we usually do not now the theoretical but only the empirical quantiles.

Furthermore, the observed instabilities in variance estimation for the Monte Carlo variance,
particularly of ALE plots, indicate room for methodological improvements. Future research may
focus on developing more better estimation procedures for decomposing variance components,
before applying these estimators to address other research questions.

@@

Further, it is important to note that our study primarily adopts a ``true to the data'' perspective
when defining feature effect errors, focusing on explaining the underlying data-generating process
rather than the model's internal representation. While this choice aligns with many practical
applications, there are also use cases for which ``true to the model'' interpretations might be 
preferred.
% also: data choice only with regard to feature effect error, not uncertainty estimation

Moreover, while our cross-validation strategy showed promising improvements over
conventional estimation on a single dataset, this approach could be enhanced further.
Future research could explore strategies that utilize more data per fold such as using the
training set in each fold for feature effect estimation (and validation only for e.g.\ perfomance estimation).
Alternatively, bootstrapping strategies similar to the approaches pursued in~\cite{molnar_relating_2023} could be
explored.

Following the observations of Loecher~\cite{loecher_debiasing_2024} that MDI and SHAP show
a bias toward high-entropy variables for Random Forests, future work might also investigate if and how
feature entropy and cardinality influence feature effect estimation.

While we aimed to incorporate real-world relevance through physics-inspired datasets, further
work is needed to fully capture the complexity of practical applications. In practice, datasets
have features with varying and also just partial effect strengths, different noise levels,
higher-order interactions, and diverse marginal distributions. Future work should focus on
developing more realistic datasets capturing these aspects, to further bridge the gap between
research, such as simulation studies, and practical applications.

@@

In the following, we provide the complete results of the ablation study on variance decomposition,
including all datasets, features, and considered models.
